Normalization by Evaluation for Sized Dependent Types

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Erase size annotations during read back.

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
      \RUp k T m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V}
    }{\RTy k T {\epiT \Size V}}
\end{gather*}

We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.  For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.

@@

The PERs $\NE$ characterizes neutral reifyable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \approx m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  \fbox{$n = n' \in \NE$}
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \approx m' &\mforall k \\
  \fbox{$d = d' \in \NF$}
    & \defas &
    \R k d v &\mand& \R k {d'} {v'} &\mand& v \approx v' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

\begin{lemma}[Closure properties of $\NE$ and $\NF$]
  \bla
  \begin{enumerate}
  \item $\lev k = \lev k \in \NE$.
  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' \in \NE$.
%  \item If $d = d' \in \NF$
  \end{enumerate}
\end{lemma}

@@

\nextcase $T \evalsto \piT A B$
\begin{gather*}
  \ru{A \in \SET[\ell] \qquad
      B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
      A \shape A_1 \qquad
      B[u] \shape B_1[u'] \mforall u = u' \in A
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
      A \shape A_2 \qquad
      B[u] \shape B_2[u'] \mforall u = u' \in A
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
  Let $d_i = \down{A_i}{u_i}$.
  By induction hypothesis, $d_1 = d_2 \in \NF$.
  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$, and
  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.

@@

\nextcase $T \evalsto \forallT B$
\begin{gather*}
  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.

@@

\begin{theorem}[Completeness of NbE]
  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \approx v'$ such that
  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
  by read back with $k = |\Gamma|$.
\end{proof}

>>>>

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{
    }{\RElim k {\ann \alpha} {\ann \infty}}
  % \ru{\RSize k \alpha b
  %   }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\R k D V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat \alpha \qquad
      \RSize k \alpha b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{% T \evalsto \up{U}n \qquad \RNe k n m
      \RUp k T m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V}
    }{\RTy k T {\epiT \Size V}}
\end{gather*}

% We write \fbox{$t \approx t'$} iff $t$ and $t'$ are syntactically equal if we ignore sizes $\ann a$ in irrelevant positions.  For example, $\suc a \tzero \approx \suc 0 \tzero$, but $\Nat a \not\approx \Nat 0$.

@@

The PERs $\NE$ characterizes neutral reifyable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  \fbox{$n = n' \in \NE$}
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
  \fbox{$d = d' \in \NF$}
    & \defas &
    \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
  \fbox{$e = e' \in \ELIM$}
    & \defas &
    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
  \fbox{$A = A' \in \TY$}
    & \defas &
    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

\begin{lemma}[Closure properties of $\NE$]
  \label{lem:closne}
  \bla
  \begin{enumerate}
  \item $\lev k = \lev k \in \NE$.
  \item If $n = n' \in \NE$ and $e = e' \in \ELIM$ then $n\,e = n'\,e' \in \NE$.
  \end{enumerate}
\end{lemma}

\begin{lemma}[Closure properties of $\ELIM$]
  \label{lem:closelim}
  \bla
  \begin{enumerate}
  \item If $d = d' \in \NF$ then $d = d' \in \ELIM$.
  \item If $\alpha \in \SIZE$ then $\alpha = \alpha \in \ELIM$.
  \item $\ann \alpha = \ann{\alpha'} \in \ELIM$.
  \item If $A = A' \in \TY$ and $d_z = d_z' \in \NF$ and $d_s = d_s' \in \NF$
        then $\case[\ell] A {d_z}{d_s} = \case[\ell]{A'}{d_z'}{d_s'} \in \ELIM$.
  \item If $D = D' \in \NF$ and $d = d' \in \NF$ then $\fix[\ell] D d = \fix[\ell]{D'}{d'} \in \ELIM$.
  \end{enumerate}
\end{lemma}

@@

\nextcase $T \evalsto \piT A B$
\begin{gather*}
  \ru{A \in \SET[\ell] \qquad
      B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
      A \shape A_1 \qquad
      B[u] \shape B_1[u'] \mforall u = u' \in A
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
      A \shape A_2 \qquad
      B[u] \shape B_2[u'] \mforall u = u' \in A
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
  Let $d_i = \down{A_i}{u_i}$.
  By induction hypothesis, $d_1 = d_2 \in \NF$.
  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$ by Lemma~\ref{lem:closne}, and
  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.

@@

\nextcase $T \evalsto \forallT B$
\begin{gather*}
  \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.

@@

\begin{theorem}[Completeness of NbE]
  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
  by read back with $k = |\Gamma|$.
\end{proof}
